GSE74548: Gene Methylation Enrichment Analysis - accounting for correlation

### 2023-09-14 figs/gmea.Rmd
refreshing hcy work

Here we will attempt another approach.
First aggregate median signal for probes belonging to a gene, followed by limma and enrichment analysis.

@@

suppressPackageStartupMessages({
  library("limma")
  library("parallel")
  library("dplyr")
  library("kableExtra")
  library("eulerr")
  library("mitch")
  library("IlluminaHumanMethylation450kmanifest")
  library("IlluminaHumanMethylation450kanno.ilmn12.hg19")
  library("tictoc")
})

>>>>

Here we will attempt functional enrichment analysis on this data, firstly using GOMeth/gsameth, and secondly
with a new method called GMEA (Gene Methylation Enrichment Analysis).
GMEA involves first aggregating median signal for probes belonging to a gene, followed by limma and subsequentenrichment analysis.

@@

suppressPackageStartupMessages({
  library("limma")
  library("parallel")
  library("dplyr")
  library("kableExtra")
  library("eulerr")
  library("mitch")
  library("IlluminaHumanMethylation450kmanifest")
  library("IlluminaHumanMethylation450kanno.ilmn12.hg19")
  library("missMethyl")
  library("tictoc")
})

### 2023-09-14 old/GSE158433/GSE158422_gmea.Rmd
refreshing hcy work

@@

hist(dm$t)
tic() ; res <- calc_sc(dm) ; toc() #32 cores 95.5
res2 <- res[[2]]

@@

hist(dm$t)
tic() ; res <- calc_sc2(dm) ; toc() #
res2 <- res[[2]]

@@

hist(dm$t)
head(dm)
tstats <- dm$t
names(tstats) <- rownames(dm)

@@

sig <- subset(fgseaRes,padj<0.05)
SIG = nrow(sig)

plot(fgseaRes$ES,-log10(fgseaRes$pval))
points(sig$ES,-log10(sig$pval),col="red")

plot(fgseaRes$size,-log10(fgseaRes$pval),log="x")
points(sig$size,-log10(sig$pval),col="red")

@@

TODO

>>>>

hist(dm$t)

@@

tic()
resw <- calc_sc(dm)
toc() #16 cores 129 sec
res2 <- resw[[2]]

@@

tic()
rest <- calc_sc2(dm)
toc() #16 cores 252 sec

res2 <- rest[[2]]

@@

head(dm)
tstats <- dm$t
names(tstats) <- rownames(dm)

@@

sig <- subset(fgseaRes,padj<0.05)
SIG = nrow(sig)
UP = nrow(subset(sig,ES>0))
DN = nrow(subset(sig,ES<0))

plot(fgseaRes$ES,-log10(fgseaRes$pval))
points(sig$ES,-log10(sig$pval),col="red")
mtext(paste(SIG,"signifiant genes,",UP,"up,",DN,"down"))

plot(fgseaRes$size,-log10(fgseaRes$pval),log="x")
points(sig$size,-log10(sig$pval),col="red")
mtext(paste(SIG,"signifiant genes,",UP,"up,",DN,"down"))

## Using CAMERA

Need to add volcano and nprobe charts here.

```{r,cam_func}

stat <- dm$t
names(stat) <- rownames(dm)

# parallel approach
CORES=detectCores()
#sets <- head(sets,100000)
sets_split <- split(sets,1:CORES)

tic()

cres <- mclapply(X=sets_split, function(mysets) {
  cameraPR(statistic=stat, index=mysets, use.ranks = FALSE, inter.gene.cor=0.01, sort = TRUE)
}, mc.cores=CORES)

cres <- do.call(rbind,cres)
rownames(cres) <- sub("\\.","@", rownames(cres))
rownames(cres) <- sapply(strsplit(rownames(cres),"@"),"[[",2)
cres <- cres[order(cres$PValue),]
g <- rownames(cres)[1]

m <- mclapply(rownames(cres) , function(g) {
  probes <- sets[[g]]
  scores <- stat[which(names(stat) %in% probes)]
  mymedian <- median(scores)
  mymean <- mean(scores)
  c(mymean,mymedian)
},mc.cores=CORES)

mdf <- do.call(rbind,m)
colnames(mdf) <- c("mean","median")
cres <- cbind(mdf,cres)
cres$FDR <- p.adjust(cres$PValue,method="fdr")
toc()

head(subset(cres,Direction=="Up"),25)
head(subset(cres,Direction=="Down"),25)

head(cres,50) %>%
  kbl(caption = "Top significant genes with CAMERA") %>%
  kable_paper("hover", full_width = F)

sig <- subset(cres,FDR<0.05)
SIG <- nrow(sig)

plot(cres$median,-log10(cres$PValue))
points(sig$median,-log10(sig$PValue),col="red")

plot(cres$NGenes,-log10(cres$PValue),log="x")
points(sig$NGenes,-log10(sig$PValue),col="red")

@@

ROAST is self contained test using rotations.

## TODO

The goal is to use a self contained test that can handle extensive correlation.

Globaltest, ANCOVA and SAM-GS

GSVA

Q2 (Tian et al. [14])
14. Tian L, Greenberg SA, Kong SW, et al. Discovering statistically significant pathways in expression profiling studies. Proc Natl Acad Sci USA. 2005;102(38):13544–9. [PMC free article] [PubMed] [Google Scholar]

FCS.SC—self-contained version of FCS (Pavlidis et al. [24])
24. Pavlidis P, Qin J, Arango V, et al. Using the gene ontology for microarray data mining: a comparison of methods and application to age effect in human prefrontal cortex. Neurochem Res. 2004;29(6):1213–22. [PubMed] [Google Scholar]

### 2023-09-14 old/GSE158433/GSE158433_integration.Rmd
refreshing hcy work

suppressPackageStartupMessages({
  library("plyr")
  library("R.utils")
  library("missMethyl")
  library("limma")
  library("DMRcate")
  library("DMRcatedata")
  library("topconfects")
  library("minfi")
  library("IlluminaHumanMethylation450kmanifest")
  library("RColorBrewer")
  library("IlluminaHumanMethylation450kanno.ilmn12.hg19")
  library("GEOquery")
  library("eulerr")
  library("plyr")
  library("gplots")
  library("reshape2")
  library("forestplot")
  library("beeswarm")
  library("RCircos")
  library("qqman")
  library("ENmix")
  library("tictoc")
  library("mitch")
  library("kableExtra")
})

source("../meth_functions.R")
CORES=detectCores()/2

@@

anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
myann <- data.frame(anno[,c("UCSC_RefGene_Name","Regulatory_Feature_Group","Islands_Name","Relation_to_Island")])
promoters <- grep("Prom",myann$Regulatory_Feature_Group)

@@

## GMEA whole gene

@@

>>>>

suppressPackageStartupMessages({
  library("plyr")
  library("R.utils")
  library("missMethyl")
  library("limma")
  library("DMRcate")
  library("DMRcatedata")
  library("topconfects")
  library("minfi")
  library("IlluminaHumanMethylation450kmanifest")
  library("RColorBrewer")
  library("IlluminaHumanMethylation450kanno.ilmn12.hg19")
  library("GEOquery")
  library("eulerr")
  library("plyr")
  library("gplots")
  library("reshape2")
  library("forestplot")
  library("beeswarm")
  library("RCircos")
  library("qqman")
  library("ENmix")
  library("tictoc")
  library("mitch")
  library("kableExtra")
  library("fgsea")
})

source("../meth_functions.R")
CORES=8

@@

anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
myann <- data.frame(anno[,c("UCSC_RefGene_Name",
  "Regulatory_Feature_Group","Islands_Name","Relation_to_Island")])
promoters <- grep("Prom",myann$Regulatory_Feature_Group)

Gene probe sets

```{r,sets}

gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
names(gp2) <- rownames(gp)
sets <- split(rep(names(gp2), lengths(gp2)), unlist(gp2))

hist(unlist(lapply(sets,length)),xlim=c(0,200),breaks=500,xlab="set size",main="probes per gene")

@@

## GMEA whole gene - wilcox test

@@

Investigate how probe number influences significance.

```{r,probebias}

plot(res2$nprobes,res2$sig,log="x",ylim=c(0,50),pch=19,cex=0.6)
points(sig$nprobes,sig$sig,col="red",pch=19,cex=0.62)
MIN = min(sig$nprobes)
LEFT = nrow(subset(res2,nprobes<MIN))
RIGHT = nrow(subset(res2,nprobes>MIN))
SIG = nrow(sig)
TOT = nrow(res2)
HEADER <- paste(TOT, "genes in total.", SIG, "with FDR<0.05.",
  RIGHT, "well covered and", LEFT, "poorly covered")
mtext(HEADER)
abline(v=MIN,lty=2)

### 2023-09-14 old/gmea_camera.Rmd
refreshing hcy work

---
title: "GSE74548: Gene Methylation Enrichment Analysis"
author: "hcy_meth meta-analysis group"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
---

@@

In this analysis we will use the principle of Gene Set Enrichment Analysis, applying it
to many probes belonging to genes.
If the probes are trending in concert, then we can make some judgement about the
enrichment of those probes.
The statistical test used is the Wilcox test, which can be applied to competitive 
or self contained test types.

@@

library("parallel")
library("dplyr")
library("kableExtra")
library("eulerr")
library("mitch")
library("IlluminaHumanMethylation450kmanifest")
library("IlluminaHumanMethylation450kanno.ilmn12.hg19")
library("tictoc")

@@

@@

@@

>>>>

---
title: "GSE74548: Gene Methylation Enrichment Analysis - accounting for correlation"
author: "hcy_meth meta-analysis group"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
---

@@

In this analysis we will use the principle of Gene Set Enrichment Analysis, applying it
to many probes belonging to genes.
If the probes are trending in concert, then we can make some judgement about the
enrichment of those probes.
The statistical test used is the CAMERA test, which is a competitive test that attempts to account for
correlation between genes in a set, or in this case probes belonging to a gene.

@@

library("parallel")
library("dplyr")
library("kableExtra")
library("eulerr")
library("mitch")
library("limma")
library("IlluminaHumanMethylation450kmanifest")
library("IlluminaHumanMethylation450kanno.ilmn12.hg19")
library("tictoc")

@@

CORES=detectCores()

@@

# make a list of probes that belong to a gene
gn <- unique(unlist(strsplit( myann$UCSC_RefGene_Name ,";")))
gnl <- strsplit( myann$UCSC_RefGene_Name ,";")
gnl <- mclapply(gnl,unique,mc.cores=CORES)
myann$UCSC_RefGene_Name <- gnl

gnl <- gnl[which(lapply(gnl,length)>0)]

if ( ! file.exists("sets.Rds") ) {

  l <- mclapply(1:nrow(myann), function(i) {
    a <- myann[i,]
    len <- length(a[[1]][[1]])
    probe <- rep(rownames(a),len)
    genes <- a[[1]][[1]]
    data.frame(genes,probe)
  },mc.cores=CORES)

  df <- do.call(rbind,l)

  sets <- mclapply(X=gn, FUN=function(g) {
    df[which(df$genes == g),2]
  } , mc.cores=CORES/2)

  names(sets) <- gn

  saveRDS(object=sets,file="sets.Rds")

} else {

  sets <- readRDS("sets.Rds")

}

@@

## Using CAMERA

```{r,cam_func}

stat <- dm$t
names(stat) <- rownames(dm)

# slow
sets2 <- head(sets,1000)
tic()
cres <- cameraPR(statistic=stat, index=sets2, use.ranks = FALSE, inter.gene.cor=0.01, sort = TRUE)
toc()

# fast
CORES=detectCores()
#sets <- head(sets,100000)
sets_split <- split(sets,1:CORES)

tic()
cres <- mclapply(X=sets_split, function(mysets) {
  cameraPR(statistic=stat, index=mysets, use.ranks = FALSE, inter.gene.cor=0.01, sort = TRUE)
}, mc.cores=CORES)
cres <- do.call(rbind,cres)
rownames(cres) <- sapply(strsplit(rownames(cres),"\\."),"[[",2)
cres <- cres[order(cres$PValue),]
g <- rownames(cres)[1]
m <- mclapply(rownames(cres) , function(g) {
  probes <- sets[[g]]
  scores <- stat[which(names(stat) %in% probes)]
  mymedian <- median(scores)
  mymean <- mean(scores)
  c(mymean,mymedian)
},mc.cores=CORES)
mdf <- do.call(rbind,m)
colnames(mdf) <- c("mean","median")
cres <- cbind(mdf,cres)
toc()

head(subset(cres,Direction=="Up"),25)
head(subset(cres,Direction=="Down"),25)

head(cres,50) %>%
  kbl(caption = "Top significant genes with CAMERA") %>%
  kable_paper("hover", full_width = F)

## CAMERA + pathway analysis

There's no ES and the p-value is biased, so not sure about this
